\pkg{evtree}: Evolutionary Learning of Globally Optimal Classification and Regression Trees in \proglang{R}

### 2011-09-26 pkg/evtree/inst/doc/evtree.Rnw
Improved almost all sections of the paper, a few FIXMEs remain.
Compilation requires deleting the cached .rda first.



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@249 255f4249-2263-4417-b1ea-8498779e27a7

@@

Classification and regression tree analysis aims at
modeling a response variable $Y$ by a vector of $P$ predictor variables
$X=(X_1,...,X_P)$. For classification trees the output variable $Y$ is
qualitative and for regression trees $Y$ is quantitative. Tree-based methods
first partition the input space $X$ into a set of $M$ rectangular regions $R_m$
($m = 1,...,M$) then fit a (typically simple) model within each region $\{Y | X \in R_m\}$,
e.g., the mean, median, or variance etc. Typically, the mode is used for
classification trees and the arithmetic mean is applied for regression trees.

In this work only binary tree models are considered, where regions are split
into two halves at a time. However, this is not a big issue in global tree
induction. As binary splits are searched globally and any multiway split can be
represented by a series of binary splits. Based on a set of split rules, binary
classification and regression trees recursively partition the input space $X$
into two regions. Samples that fulfill the criterion of a split rule propagate
down into the descendant left node and the other instances into the right node.
Split points of a tree are termed \textit{internal nodes} and nodes without
successors are called \textit{terminal nodes}. A binary tree with $M$ terminal
nodes has $M-1$ internal nodes. Generally, the maximum number of terminal nodes
$M_{max}=\infty$. In practice, $M_{max}$ is restricted by the maximum tree depth
$D_{max}$ as $M_{max} = 2^{D_{max}}-1$. When the trees are constrained to have a
minimum number of observations in the internal and/or terminal nodes, $M_{max}$
is also restricted by the finiteness of the training data.

A binary tree model can be uniquely defined by the positions of the internal
nodes, being a subset of $\{1,...,M_{max}-1\}$, and the corresponding split
rules. Let the positions of the internal nodes of a tree with $M-1$ terminal
nodes be indicated by $n_1,...,n_{M-1}$. The positions $\{n_1,...,n_{M-1}\} \in
\{1,...,M_{max}-1\}$ and $n_i < n_j \Leftrightarrow i < j$. Let the variable
selected for splitting, at a new position $r \in n_1,...,n_{M_{max}-1}$, be
denoted as $v_r$, where $v_r=1,...,P$. Depending on the domain of $X_{v_r}$, a
split point or a set of categories has to be selected, which we denote as $s_r$.
If $X_{v_r}$ is numerical or ordinal, a binary split rule is defined as $X_{v_r}
< s_r$. If $X_{v_r}$ has $u$ unique values, then there are $u-1$ possible split
points and $s_r=1,...,u-1$. In the case of a nominal split variable $X_{v_r}$, a
split rule is defined as $X_{v_r} \in s_r$, where $s_r \subset \{1,...,K\}$ and
$s_r \neq \{\emptyset\}$. If $X_{v_r}$ has $K$ categories there are $2^{K-1}-1$
possible splits.

@@

As done by \cite{breiman1984cart}, let the complexity of a tree be measured by a
function of the number of terminal nodes, without further considering the depth
or the shape of trees. Then the overall parameter space $\Theta$ can be written
as $\Theta =\bigcup_{M=1}^{M_{max}} \Theta_{M}$, where $\Theta_M$ denotes the
parameter space of models with $M$ terminal nodes. Let $\theta$ be a generic
element of $\Theta$, then $\theta$ is fully defined by:
%
\begin{equation}
	 \theta=(v_{n_1}, s_{n_1},...,v_{n_{M-1}},s_{n_{M-1}} ).
	\label{treeRep}
\end{equation}
%
The overall goal of classification and regression tree analysis can be defined
as finding a set of parameters $\hat{\theta}$, such that
%
\begin{equation}
	\hat{\theta} = \argmin_{\theta \in \Theta} \; \mathit{loss}(Y, f(X,\theta)) + \mathit{comp}(\theta).
	\label{generalcostfunction}
\end{equation}
%
Based on the input data $X$ and the parameters $\theta$, $f(X,\theta)$ gives
predictions for the output $Y$. $loss$ is a suitable loss function for the
domain of $Y$. Commonly, the misclassification error for classification
problems, or the squared error for regression problems, is used. $comp$
quantifies the complexity of trees and penalizes for more complex models in the
tree selection process. In CART, the complexity is a non-negative function of
the terminal-nodes. Note that finding $\hat{\theta}$ requires a search over all
$\Theta_m$.

The parameter space $\Theta$ becomes large for already medium sized problems and
a complete search for larger problems is computationally intractable. In fact,
\cite{hyafil1976cob} showed that building optimal binary decision trees, such
that the expected number of tests required to classify an unknown sample is
minimized, is NP-complete. \cite{zantema2000finding} proved that finding a
decision tree of minimal size that is decision equivalent to a given decision
tree is an NP-hard problem. As a consequence the search space is usually limited
by heuristics.

@@

Instead of searching all combination of $\Theta$ simultaneously, recursive
partitioning algorithms only consider one split at a time. At each internal node
$r \in \{n_1,..., n_{M-1}\}$, the split variable $X_{v_r}$ and the corresponding
split point $s_r$ is selected such that it minimizes the impurity in the two
resulting terminal nodes. Let the parameters of a tree, constructed in
$i$-steps, be denoted as $\tilde{\theta}_{i}$ and $\tilde{\theta}_{0} =
(\emptyset)$. Then, for a tree with parameters $\tilde{\theta}_{i}$, the
position of the new internal node $r$, as well as the corresponding split
variable $v_r$ and split point $s_r$, are chosen such that:
%
\begin{equation}
		\min_{v_r, s_r} \; \mathit{loss}(Y, f(X, \tilde{\theta}_{i-1} \cup (v_r, s_r) ))
	\label{eq:whichsplit}
\end{equation}
%
CART's standard measures for node impurity are the Gini index, for
classification problems and the mean squared error, for regression problems.
After the selection of the $i^{th}$ split $\tilde{\theta}_{i}$ is given by
$\tilde{\theta}_{i} = \tilde{\theta}_{i-1}  \cup (\hat{v}_r, \hat{s}_r)$. This
process of recursively selecting and executing splits continues until the
homogeneity of the nodes cannot be further improved. Additionally, often
additional stopping criteria are defined, e.g. minimum sample sizes in the
internal nodes and terminal nodes.

In CART, and many other recursive tree methods, the tree growing step is
followed by a pruning step. The idea behind pruning is to first grow a large
tree without or with little restriction on the magnitude of impurity reduction.
Starting from the bottom of the tree pruning then removes worthless splits. The
method proposed by Breiman \textit{et al.} \cite[Chapter 2]{breiman1984cart} and
used by CART is called \textit{cost-complexity pruning}. Cost-complexity pruning
iteratively prunes the internal node that:
%
\begin{equation}
		\min_{r=1,...,n_{M-1}} \; \mathit{loss}(Y, f(X,\tilde{\theta}_i \setminus (v_r, s_r )))  + \mathit{comp}(\tilde{\theta}_i \setminus (v_r, s_r)).
	\label{eq:pruning}
\end{equation}
%
The subsequent solution $\tilde{\theta}_{i+j+1} = \tilde{\theta}_{i+j} \setminus
(\hat{v}_r, \hat{s}_r)$, where $j$ is the number of completed pruning steps.

For nontrivial problems, recursive partitioning methods only search a small
fraction of the global search space $(v_1, s_1, ..., v_{M_{max}-1},
s_{M_{max}-1} )$. They only search each $(v_r, s_r)$ once, and independently of
the other split rules. The selection of $v_r$ and $s_r$ at each step $i$
requires a search of the locally optimal split over $M=i$ terminal nodes. The
pruning step requires $j$ selections of an internal node $r$.

In contrast to CART and C4.5, conditional inference trees utilize an unbiased
variable selection approach that is based on permutation testing. Bias in the
split variable selection process arises from a preference towards variables with
many potential split points. The conditional inference framework overcomes this
problem by separating the variable and split point selection. Although this
distinction influences the selection of split rules, conditional inference tree
solve the same minimization problem (Equation~\ref{eq:whichsplit}). The
\code{ctree} implementation of conditional inference trees does not employ
pruning, but uses a statistical stopping criterion.

@@

A very simple example that illustrates the limitation of recursive partitioning
methods is depicted in Figure~\ref{fig:illustration}. The example only contains
two independent variables and can be solved with only three splits that
partition the input space into four regions. As expected the recursive
partitioning methods \code{rpart} and \code{ctree} fail to find any split at
all, as the impurity of the resulting subsets cannot be reduced by the first
split. For methods that explore $\Theta$ in a more global fashion it is
straightforward to find an optimal solution to this problem. One solution, which
is the tree constructed by \code{evtree} is given by: 
%
<<chess22, echo=FALSE>>=
X1 <- rep(seq(0.25, 1.75, 0.5), each = 4)
X2 <- rep(seq(0.25, 1.75, 0.5), 4)
Y <- rep(1, 16)
Y[(X1 < 1 & X2 < 1) | (X1 > 1 & X2 > 1)] <- 2
Y <- factor(Y, labels = c("O", "X"))
chess22 <- data.frame(Y, X1, X2)
print(evtree(Y ~ ., data = chess22, minbucket = 1, minsplit = 2))
@

@@

All instances all are classified correctly. Each of the terminal nodes~3 and~7 contain four instances of the class \code{X}. Four instances of class
\code{O} are assigned to each of the terminal nodes~4 and~6.

@@

When compared with the described forward stepwise search, a less greedy approach is to calculate the effects of the choice of the split rule deeper down in the tree. In this way optimal trees can be found for simple problems. However, feature selection at a node of greedily induced trees has complexity  $O(PN)$. Through a search of the optimal split up to $d-1$ levels the complexity increases to $O(P^dN^d)$ \citep{papagelis2001bdt}. Therefore, for more difficult problems the search space at each node must be limited by the depth of the search. Additionally, the search space can be reduced, limiting the number of possible split points of the selected variables and/or other heuristics. These methods are known as \textit{look-ahead algorithms} \citep{esmeir2007ald}. Another class of algorithms is given by stochastic optimization methods. Given an initial tree, these methods seek improved solutions through stochastic changes to the tree structure. Besides evolutionary algorithms \citep{koza1991concept}, \textit{Bayesian CART} \citep{denison1998bca} and \textit{simulated annealing} \citep{sutton1992ict} were used successfully to solve difficult classification and regression tree problems. \cite{koza1991concept} first formulated the concept of using evolutionary algorithms as a stochastic optimization method to build classification and regression trees. \cite{papagelis2001bdt} presented a classification tree algorithm and provided results on several datasets from the UCI machine learning repository \citep{Frank+Asuncion:2010}. A method for the construction of regression trees via evolutionary algorithms was introduced by \cite{fan2005rta}.  \cite{cantu2003inducing} used an evolutionary algorithm to induce oblique classification trees. \cite{gray2008cta} introduced a classification tree algorithm that allows univariate splits, but also splits defined by linear combinations of two or three variables.

@@

The pseudocode for the general scheme of an evolutionary algorithm is depicted in Table~\ref{pc:genericEA}. With the description of these components together with the definition of individuals, the most important aspects of an evolutionary algorithm are specified. The representation of an individual is given by a binary tree structure, which is fully defined with the definition of $\theta$ in Equation~\ref{treeRep}. This section continues with the description of the remaining components following the general outline displayed in Table~\ref{pc:genericEA}.

\begin{table}[t!]
\centering
\hrulefill
\begin{enumerate}
  \item Initialize the population.
  \item Evaluate each individual.
  \item While(termination condition is not satisfied) do:
  \begin{itemize}
    \item[a.] Select parents.
    \item[b.] Alter selected individuals via variation operators.
    \item[c.] Evaluate new solutions.
    \item[d.] Select survivors for the next generation.
  \end{itemize}
\end{enumerate}
\hrulefill
\caption{Pseudocode of the general schema of an evolutionary algorithm.} 
\label{pc:genericEA}
\end{table}

This section describes the general functioning of the \code{evtree} algorithm. Implementation details as user-specified parameters, including their default values and their possible ranges, are discussed in the subsequent section. These implementation details also include parameters that specify a minimum number of observation that are required in each internal node and  terminal node of a tree. In the following, the term \textit{invalid node} refers to a node that do not contain enough observations.

@@

Each tree of the population is initialized with a valid, randomly generated, split rule in the root node. First, $v_1$ is selected with uniform probability from $1,...,P$. Second, a split point $s_1$ is selected. If $X_{v_1}$ is numeric or ordinal with $u$ unique values, a split point $s_1$ is selected with uniform probability from the $u-1$ possible spit points of $X_{v_1}$. If $X_{v_1}$ is nominal and has $K$ categories, each $k=1,...,K$ has a probability of $50\%$ to be assigned to the left or the right terminal node. In cases where all $k$ are allocated to the same terminal node, the assignment of one category is flipped to the other terminal node. If this procedure results in a non-valid split rule, the two steps of random split variable selection and split point selection are repeated. With the definition of $r=1$ and the selection of $v_1$ and $s_1$, the initialized tree is fully described by $\tilde{\theta}_{1}= (v_1, s_1)$.

@@

The split rule of the new internal node is chosen as follows. First, $v_r$ is selected with uniform probability from $1,...,P$. Second, a split point $s_r$ is selected. If $X_{v_r}$ is numeric or ordinal, a split point $s_r$ is selected with uniform probability from the $u-1$ possible spit points of $X_{v_r}$. If $X_{v_r}$ is nominal, each category of $X_{v_r}$, has a probability of $50\%$ to be assigned to the left or the right terminal node. In cases where all $k$ are allocated to the same terminal node, one of them is changed to the other terminal node. If this procedure results in a non-valid split rule, the two steps of random split variable selection and split point selection are repeated for a maximum of $P$ iterations.

In cases where no valid split rule can be assigned to the internal node at position $r$, the search for a valid split rule is carried out on another randomly selected terminal node. If, after $10$ attempts, no valid split rule can be found, then $\tilde{\theta}_{i+1} = \tilde{\theta}_{i}$. Otherwise, the set of parameters in iteration $i+1$ are given by $\tilde{\theta}_{i+1}= \tilde{\theta}_i \cup (v_{r},s_{r})$.

\subsubsection{Prune}
\textit{Prune} chooses a random internal node $r$, where $r>1$, which has two terminal nodes as successors and prunes it into a terminal node. The tree's parameters at iteration $i+1$ are $\tilde{\theta}_{i+1}=\tilde{\theta}_{i} \setminus (v_{r},s_{r})$. If $\tilde{\theta}_{i}$ only comprises one internal node, i.e., the root node, than $\tilde{\theta}_{i+1} = \tilde{\theta}_{i}$.

\subsubsection{Major split rule mutation}
\textit{Major split rule mutation} selects a random internal node $r$ and changes the split rule, defined by the corresponding split variable $v_r$, and the split point $s_r$. With a probability of $50\%$, a value from the range $1,...,P$ is assigned to $v_r$. Otherwise $v_r$ remains unchanged and only $s_r$ is modified. Again, depending on the domain of $X_{v_r}$, either a random split point from the range of possible values of $X_{v_r}$ is selected, or a non-empty set of categories is assigned to each of the two terminal nodes. If the split rule at $r$ becomes invalid, the mutation operation is reversed and the procedure, starting with the selection of $r$, is repeated for a maximum of $3$ attempts. Subsequent nodes that become invalid are pruned.

@@

\textit{minor split rule mutation} is similar to the \textit{major split rule mutation} operator, but it does not alter $v_r$ and only changes the split point $s_r$ by a minor degree.
If $X_{v_r}$ is numerical or ordinal the split point $s_r$ is changed by a non-zero number of unique values of $X_{v_r}$. In cases where $X_{v_r}$ has less then $20$ unique values, the split point is change to to the next larger, or the next lower, unique value of $X_{v_r}$. Otherwise, $s_r$ is randomly shifted by a number of unique values that is not larger than $10\%$ of the range of unique values of $X_{v_r}$. If $X_{v_r}$ is a nominal variable, with less than $20$ categories, one of the categories is randomly modified. Otherwise, at least one and at most $10\%$ of the variable's categories are changed. In cases where subsequent nodes become invalid, further split points are searched that preserve the tree's topology. After five non-successful attempts at finding a topology preserving split point, the non-valid nodes are pruned.

@@

\textit{Crossover} exchanges, randomly selected, subtrees between two trees. Let $\tilde{\theta}_{i}^1$ and $\tilde{\theta}_{i}^2$ be the parameters of the trees that are chosen for crossover. First, an internal node is selected randomly from each of $\tilde{\theta}_{i}^1$ and $\tilde{\theta}_{i}^2$, which we denote as $n_k^1$ and $n_k^2$. Let $sub1(\tilde{\theta}_i^j, n^j_k)$ denote the subtrees rooted by $n_k^j$, where $j=1,2$. Then, $sub1(\tilde{\theta}_i^j)$ contains the parameters $(v_{n^j_k}, s_{n^j_k},  v_{n^j_{k1}}, s_{n^j_{k1}}, ...,v_{n^j_{km}}, s_{n^j_{km}})$, where $(n^j_{k1}, ..., n^j_{km})$ are the set of descendent node positions of $n^j_k$. Now, the two selected subtrees can be written as $sub1(\tilde{\theta}_{i}^1, n^1_k) $ and $sub1(\tilde{\theta}_{i}^2, n^2_k)$. Furthermore, $sub2(\tilde{\theta}_{i}^1, n^1_k) = \tilde{\theta}_{i}^1 \setminus sub1(\tilde{\theta}_{i}^1, n^1_k)$ and $sub2(\tilde{\theta}_{i}^2, n^2_k) = \tilde{\theta}_{i}^2 \setminus sub1(\tilde{\theta}_{i}^2, n^2_k)$.

After the application of the \textit{crossover} operator, the trees' parameter spaces are given by $\tilde{\theta}_{i+1}^1= sub2(\tilde{\theta}_{i}^1, n^1_k) \cup sub1(\tilde{\theta}_{i}^2, n^2_k) \setminus (v_{m^2_{1}}, s_{m^2_{1}}, ..., v_{m^2_{f}}, s_{m^2_{f}})$ and $\tilde{\theta}_{i+1}^2= sub2(\tilde{\theta}_{i}^2, n^2_k) \cup sub1(\tilde{\theta}_{i}^1, n^1_k) \setminus (v_{m^1_{1}}, s_{m^1_{1}}, ..., v_{m^1_{f}}, s_{m^1_{f}})$. Here, $(v_{m^1_{1}}, s_{m^1_{1}}, ..., v_{m^1_{f}}, s_{m^1_{f}}) \subset sub1(\tilde{\theta}_i^1, n^1_k)$ and $(v_{m^2_{1}}, s_{m^2_{1}}, ..., v_{m^2_{f}}, s_{m^2_{f}}) \subset sub1(\tilde{\theta}_i^2, n^2_k)$ are the parameters of the invalid nodes.

@@

The evaluation function represents the requirements the population should adapt to. In general, these requirements are formulated by Equation~\ref{generalcostfunction}. A suitable evaluation function for classification and regression trees minimizes the models' accuracy on the training data, and the models' complexity. This subsection describes the currently implemented choices of evaluation functions for classification and for regression.

>>>>

%% for internal use
\newcommand{\fixme}[1]{\emph{\marginpar{FIXME} (#1)}}
\newcommand{\readme}[1]{\emph{\marginpar{README} (#1)}}

@@

Classification and regression tree analysis aims at
modeling a response variable $Y$ by a vector of $P$ predictor variables
$X=(X_1,...,X_P)$ where for classification trees $Y$ is
qualitative and for regression trees $Y$ is quantitative. Tree-based methods
first partition the input space $X$ into a set of $M$ rectangular regions $R_m$
($m = 1,...,M$) then fit a (typically simple) model within each region $\{Y | X \in R_m\}$,
e.g., the mean, median, or variance etc. Typically, the mode is used for
classification trees and the arithmetic mean is applied for regression trees.

To show why forward-search recursive partitioning algorithms typically
lead to globablly suboptimal solutions, their parameter spaces and optimization
problems are presented and contrasted in a unified notation. Although all
arguments hold more generally, only binary tree models with some maximum number
of splits $M_{\max}$ are considered. Both restrictions make the notation somewhat
simpler while not really restricting the problem: (a)~Multiway splits are equivalent
to a sequence of binary splits in predictions and number of resulting subsamples.
(b)~The maximal size of the tree is always limited by the number of observations
in the learning sample.

In the following, a binary tree model with $M$ terminal nodes (which consequently has
$M-1$ internal splits) is denoted by
%
\begin{equation}
  \theta ~=~ (v_{n_1}, s_{n_1},...,v_{n_{M-1}},s_{n_{M-1}} ), \label{eq:tree}
\end{equation}
%
where the $n_r \in \{1,...,M_{\max}-1\}$ are the positions of the
internal \emph{nodes}, $v_r \in \{1, \dots ,P\}$ the associated splitting
\emph{variables}, and $s_r$ the associated \emph{split} rules ($r = 1, \dots, M-1$).
Depending on the domain of $X_{v_r}$, the split rule $s_r$ contains either a cutoff
(for ordered and numeric variables) or a a nonempty subset of $\{1, \dots, c\}$
(for a categorical variable with $c$~levels), determining which observations are
sent to the first or second subsample. In the former case, there are $u-1$ possible
split rules if $X_{v_r}$ takes $u$ distinct values; and in the latter case, there
are $2^{c-1} - 1$ possible splits. Thus, the product of all of these combinations
forms all potential elements $\theta$ from $\Theta_M$, the space of conceivable
trees with $M$ terminal nodes. The overall parameter space is then 
$\Theta =\bigcup_{M=1}^{M_{\max}} \Theta_{M}$ (which in practice is often reduced
by exluding elements $\theta$ resulting in too small subsamples etc.).

Finally, $f(X,\theta)$ denotes the prediction function based on all explanatory
variables $X$ and the chosen tree structure $\theta$ from Equation~\ref{eq:tree}.
As pointed out above, this is typically constructed using the means or modes in the respective
partitions of the learning sample.

@@

As done by \cite{breiman1984cart}, let the complexity of a tree be measured by a
function of the number of terminal nodes, without further considering the depth
or the shape of trees. The goal is then to find that classification and regression tree
which optimizes some tradeoff between predicition performance and complexity:
%
\begin{equation} \label{eq:globally}
  \hat{\theta} ~=~ \argmin_{\theta \in \Theta} \; \mathrm{loss}\{Y, f(X,\theta)\} ~+~ \mathrm{comp}(\theta).
\end{equation}
%
where $\mathrm{loss}(\cdot, \cdot)$ is a suitable loss function for the
domain of $Y$; typically, the misclassification rate MC 
and the mean squared error MSE are employed for classification and regression,
respectively. The function $\mathrm{comp}(\cdot)$ is a function that is monotonically
non-decreasing in the number of terminal nodes $M$ of the tree $\theta$, thus penalizing
more complex models in the tree selection selection process.
Note that finding $\hat{\theta}$ requires a search over all $\Theta_M$.

The parameter space $\Theta$ becomes large for already medium sized problems and
a complete search for larger problems is computationally intractable. In fact,
\cite{hyafil1976cob} showed that building optimal binary decision trees, such
that the expected number of splits required to classify an unknown sample is
minimized, is NP-complete. \cite{zantema2000finding} proved that finding a
decision tree of minimal size that is decision-equivalent to a given decision
tree is also NP-hard. As a consequence the search space is usually limited
by heuristics.

@@

Instead of searching all combinations in $\Theta$ simultaneously, recursive
partitioning algorithms only consider one split at a time. At each internal node
$r \in \{n_1,..., n_{M-1}\}$, the split variable $v_r$ and the corresponding
split point $s_r$ are selected to locally minimize the loss function.
Starting with an empty tree $\theta_0 = (\emptyset)$, the tree is first grown 
recursively for and subsequently \emph{pruned} to satisfy the complexity tradeoff:
%
\begin{eqnarray} \label{eq:locally}
  \bar \theta_r
    & = & \argmin_{ \theta = \theta_{r-1} \cup (v_r, s_r)} \; \mathrm{loss}\{Y, f(X, \theta)\} 
    \qquad (r = 1, \dots, M_{\max} -1), \\
  \bar \theta
    & = & \argmin_{\bar \theta_r} \; \mathrm{loss}\{Y, f(X, \bar \theta_r)\} ~+~ \mathrm{comp}(\bar \theta_r).
  \label{eq:pruning}
\end{eqnarray}
%
For nontrivial problems, forward-search recursive partitioning methods only search a small
fraction of the global search space $(v_1, s_1, \dots, v_{M_{\max}-1},
s_{M_{\max}-1} )$. They only search each $(v_r, s_r)$ once, and independently of
the subsequent split rules, hence typically leading to a globally suboptimal solution
$\bar \theta$.

Note that the notation above uses an exhaustive search for the $r$-th split,
jointly over $(v_r, s_r)$, as is employed in CART or C4.5. So-called \emph{unbiased}
recursive partitioning techniques modify this search by first seleting the variable $v_r$ using
statistical significance tests and subsequently selecting the optimal split $s_r$ for that
particular variable. This approach is used in conditional inference trees
\citep[see][for references to other algorithms]{hothorn2006urp} and avoids selecting
variables with many potential splits more often than those with fewer potential
splits.

@@

A very simple example that illustrates the limitation of forward-search recursive partitioning
methods is depicted in Figure~\ref{fig:illustration}. The example only contains
two independent variables and can be solved with three splits that
partition the input space into four regions. As expected the recursive
partitioning methods \code{rpart} and \code{ctree} fail to find any split at
all, as the loss function on the resulting subsets cannot be reduced by the first
split. For methods that explore $\Theta$ in a more global fashion it is
straightforward to find an optimal solution to this problem. One solution
is the tree constructed by \code{evtree}: 
%
<<chess22, echo=FALSE>>=
X1 <- rep(seq(0.25, 1.75, 0.5), each = 4)
X2 <- rep(seq(0.25, 1.75, 0.5), 4)
Y <- rep(1, 16)
Y[(X1 < 1 & X2 < 1) | (X1 > 1 & X2 > 1)] <- 2
Y <- factor(Y, labels = c("O", "X"))
chess22 <- data.frame(Y, X1, X2)
print(evtree(Y ~ ., data = chess22, minbucket = 1, minsplit = 2))
@

@@

All instances are classified correctly. Each of the terminal nodes~3 and~7 contain four instances of the class \code{X}. Four instances of class
\code{O} are assigned to each of the terminal nodes~4 and~6.

@@

When compared with the described forward stepwise search, a less greedy approach
is to calculate the effects of the split rules deeper down in the
tree. In this way optimal trees can be found for simple problems. However,
split selection at a given node in Equation~\ref{eq:locally} has complexity  $O(PN)$
(if all $P$ variables are numeric/ordered with $N$ distinct values).
Through a global search up to $M-1$ levels (\ref{eq:globally}) the complexity
increases to $O(P^M N^M)$ \citep{papagelis2001bdt}. One conceivable compromise
between these two extremes is to look ahead $d$~steps with $1 < d < M$
\citep[see e.g.,][]{esmeir2007ald}, also yielding a locally optimal tree but
less constrained than that from a 1-step-ahead search.
Another class of algorithms is given by stochastic optimization methods that,
given an initial tree, seek improved solutions through stochastic changes to
the tree structure. Thus, these algorithms try to explore the full parameter
space $\Theta$ but cannot be guaranteed to find the globally optimal solution
but only an approximation thereof. Besides evolutionary algorithms
\citep{koza1991concept}, \textit{Bayesian CART} \citep{denison1998bca} and
\textit{simulated annealing} \citep{sutton1992ict} were used successfully to
solve difficult classification and regression tree problems.
\cite{koza1991concept} first formulated the concept of using evolutionary
algorithms as a stochastic optimization method to build classification and
regression trees. \cite{papagelis2001bdt} presented a classification tree
algorithm and provided results on several datasets from the UCI machine learning
repository \citep{Frank+Asuncion:2010}. Another method for the construction of
classification and regression trees via evolutionary algorithms was introduced by
\cite{gray2008cta} and \cite{fan2005rta}, respectively.
\cite{cantu2003inducing} used an evolutionary algorithm to induce so-called oblique
classification trees.

@@

\begin{table}[b!]
\centering
\hrulefill
\begin{enumerate}
  \item Initialize the population.
  \item Evaluate each individual.
  \item While(termination condition is not satisfied) do:
  \begin{itemize}
    \item[a.] Select parents.
    \item[b.] Alter selected individuals via variation operators.
    \item[c.] Evaluate new solutions.
    \item[d.] Select survivors for the next generation.
  \end{itemize}
\end{enumerate}
\hrulefill
\caption{Pseudocode of the general evolutionary algorithm.} 
\label{pc:genericEA}
\end{table}

The pseudocode for the general evolutionary algorithm is provided
in Table~\ref{pc:genericEA}. In the context of classification and regression trees,
all \emph{individuals} from the population (of some given size) are $\theta$s as defined in 
Equation~\ref{eq:tree}. The details of their evolutionary selection is
given below following the general outline displayed in Table~\ref{pc:genericEA}.

As pointed out in Section~\ref{got}, some elements $\theta \in \Theta$ are typically
excluded in practice to satisfy minimal subsample size requirements.
In the following, the term \emph{invalid node} refers to such excluded cases,
not meeting sample size restrictions.

@@

Each tree of the population is initialized with a valid, randomly generated,
split rule in the root node. First, $v_1$ is selected with uniform probability
from $1,...,P$. Second, a split point $s_1$ is selected. If $X_{v_1}$ is numeric
or ordinal with $u$ unique values, a split point $s_1$ is selected with uniform
probability from the $u-1$ possible spit points of $X_{v_1}$. If $X_{v_1}$ is
nominal and has $c$ categories, each $k=1,...,c$ has a probability of $50\%$ to
be assigned to the left or the right daughter node. In cases where all $k$ are
allocated to the same terminal node, the assignment of one category is flipped
to the other terminal node. If this procedure results in a non-valid split rule,
the two steps of random split variable selection and split point selection are
repeated. With the definition of $r=1$ and the selection of $v_1$ and $s_1$, the
initialization is complete and each individual of the population of trees is of
type $\tilde{\theta}_{1}= (v_1, s_1)$.

@@

The search for a valid split rule is conducted as in see Section~\ref{init}
for a maximum of $P$ iterations. 
In cases where no valid split rule can be assigned to the internal node at
position $r$, the search for a valid split rule is carried out on another
randomly selected terminal node. If, after 10~attempts, no valid split rule
can be found, then $\tilde{\theta}_{i+1} = \tilde{\theta}_{i}$. Otherwise, the
set of parameters in iteration $i+1$ are given by $\tilde{\theta}_{i+1}=
\tilde{\theta}_i \cup (v_{r},s_{r})$.

\subsubsection{Prune}

\textit{Prune} chooses a random internal node $r$, where $r>1$, which has two
terminal nodes as successors and prunes it into a terminal node. The tree's
parameters at iteration $i+1$ are $\tilde{\theta}_{i+1}=\tilde{\theta}_{i}
\setminus (v_{r},s_{r})$. If $\tilde{\theta}_{i}$ only comprises one internal
node, i.e., the root node, then $\tilde{\theta}_{i+1} = \tilde{\theta}_{i}$.

\subsubsection{Major split rule mutation}

\textit{Major split rule mutation} selects a random internal node $r$ and
changes the split rule, defined by the corresponding split variable $v_r$, and
the split point $s_r$. With a probability of $50\%$, a value from the range
$1,...,P$ is assigned to $v_r$. Otherwise $v_r$ remains unchanged and only $s_r$
is modified. Again, depending on the domain of $X_{v_r}$, either a random split
point from the range of possible values of $X_{v_r}$ is selected, or a non-empty
set of categories is assigned to each of the two terminal nodes. If the split
rule at $r$ becomes invalid, the mutation operation is reversed and the
procedure, starting with the selection of $r$, is repeated for a maximum of
3~attempts. Subsequent nodes that become invalid are pruned.

@@

\textit{Minor split rule mutation} is similar to the \textit{major split rule
mutation} operator, but it does not alter $v_r$ and only changes the split point
$s_r$ by a minor degree. If $X_{v_r}$ is numerical or ordinal the split point
$s_r$ is changed by a non-zero number of unique values of $X_{v_r}$. In cases
where $X_{v_r}$ has less then $20$ unique values, the split point is change to
to the next larger, or the next lower, unique value of $X_{v_r}$. Otherwise,
$s_r$ is randomly shifted by a number of unique values that is not larger than
$10\%$ of the range of unique values of $X_{v_r}$. If $X_{v_r}$ is a nominal
variable, with less than $20$ categories, one of the categories is randomly
modified. Otherwise, at least one and at most $10\%$ of the variable's
categories are changed. In cases where subsequent nodes become invalid, further
split points are searched that preserve the tree's topology. After five
non-successful attempts at finding a topology preserving split point, the
non-valid nodes are pruned.

@@

\fixme{I really had problems following this section. The sub-operator is useful
but is it possible to introduce this with fewer sub- and super-scripts? It makes
it very challenging to read.}

\textit{Crossover} exchanges, randomly selected, subtrees between two trees. Let
$\tilde{\theta}_{i}^1$ and $\tilde{\theta}_{i}^2$ be the parameters of the trees
that are chosen for crossover. First, an internal node is selected randomly from
each of $\tilde{\theta}_{i}^1$ and $\tilde{\theta}_{i}^2$, which we denote as
$n_k^1$ and $n_k^2$. Let $\mathrm{sub}1(\tilde{\theta}_i^j, n^j_k)$ denote the subtrees
rooted by $n_k^j$ ($j=1,2$). Then, $\mathrm{sub}1(\tilde{\theta}_i^j) =
(v_{n^j_k}, s_{n^j_k},  v_{n^j_{k1}}, s_{n^j_{k1}}, \dots ,v_{n^j_{km}}, s_{n^j_{km}})$, where $(n^j_{k1}, ..., n^j_{km})$ are the set
of descendent node positions of $n^j_k$. Now, the two selected subtrees can be
written as $\mathrm{sub}1(\tilde{\theta}_{i}^1, n^1_k) $ and $\mathrm{sub}1(\tilde{\theta}_{i}^2,
n^2_k)$. Furthermore, $\mathrm{sub}2(\tilde{\theta}_{i}^1, n^1_k) = \tilde{\theta}_{i}^1
\setminus \mathrm{sub}1(\tilde{\theta}_{i}^1, n^1_k)$ and $\mathrm{sub}2(\tilde{\theta}_{i}^2,
n^2_k) = \tilde{\theta}_{i}^2 \setminus \mathrm{sub}1(\tilde{\theta}_{i}^2, n^2_k)$.

After the application of the \textit{crossover} operator, the trees' parameter
spaces are given by $\tilde{\theta}_{i+1}^1= \mathrm{sub}2(\tilde{\theta}_{i}^1, n^1_k)
\cup \mathrm{sub}1(\tilde{\theta}_{i}^2, n^2_k) \setminus (v_{m^2_{1}}, s_{m^2_{1}}, ...,
v_{m^2_{f}}, s_{m^2_{f}})$ and $\tilde{\theta}_{i+1}^2=
\mathrm{sub}2(\tilde{\theta}_{i}^2, n^2_k) \cup \mathrm{sub}1(\tilde{\theta}_{i}^1, n^1_k)
\setminus (v_{m^1_{1}}, s_{m^1_{1}}, ..., v_{m^1_{f}}, s_{m^1_{f}})$. Here,
$(v_{m^1_{1}}, s_{m^1_{1}}, ..., v_{m^1_{f}}, s_{m^1_{f}}) \subset
\mathrm{sub}1(\tilde{\theta}_i^1, n^1_k)$ and $(v_{m^2_{1}}, s_{m^2_{1}}, ...,
v_{m^2_{f}}, s_{m^2_{f}}) \subset \mathrm{sub}1(\tilde{\theta}_i^2, n^2_k)$ are the
parameters of the invalid nodes.

@@

The evaluation function represents the requirements the population should adapt
to. In general, these requirements are formulated by Equation~\ref{eq:globally}.
A suitable evaluation function for classification and regression trees minimizes
the models' accuracy on the training data, and the models' complexity. This
subsection describes the currently implemented choices of evaluation functions
for classification and for regression.
